\documentclass[11pt,a4paper]{article}
\input{preamble.tex}
\input{commands.tex}

\title{\sffamily \bfseries A variational derivation of the field equations of an action-dependent
	Einstein-Hilbert Lagrangian}
\author{\sffamily 
 $^a$Jordi Gaset
 \thanks{\texttt{jordi.gaset@unir.ne}t\,({ORCID}:\,0000-0001-8796-3149).} ,
$^b$Arnau Mas Dorca
\thanks{\texttt{arnau.mas@physik.uni-muenchen.de}\,({ORCID}:\,0000-0003-0532-0938).}.
\\[1ex]
\normalsize\itshape\sffamily 
$^a$Escuela Superior de Ingeniería y Tecnología, Universidad Internacional de La Rioja, Spain.
\\[1ex]
\normalsize\itshape\sffamily 
$^b$ Fakultät für Physik, Ludwig-Maximilians-Universität München
}
\date{\sffamily \small June 23, 2022}

\begin{document}
\maketitle

\selectlanguage{english}

\begin{abstract}
We derive the equations of motion of an action-dependent version of the Einstein-Hilbert Lagrangian, as a specific instance of the Herglotz variational problem. Action-dependent Lagrangians lead to dissipative dynamics, which cannot be obtained with the standard method of Lagrangian field theory. First-order theories of this kind are relatively well understood, but examples of singular or higher-order action-dependent field theories are scarce. This work constitutes an example of such a theory. By casting the problem in clear geometric terms we are able to obtain a Lorentz invariant set of equations, which contrasts with previous attempts. 
\end{abstract}

\newpage
	{\small \sffamily \tableofcontents}
\newpage

\section{Background and motivation}
\subfile{chapter-1}

\section{The Herglotz variational problem}\label{ch:herglotz}
\subfile{chapter-2}

\section{Action-dependent Einstein gravity}\label{ch:einstein}
\subfile{chapter-3}

\section{Significance of the equations}\label{ch:significance}
\subfile{chapter-4}

\section{Conclusions}\label{ch:conclusions}
\subfile{chapter-5}


\setquotestyle{english}
\printbibliography

\end{document}
